BOLD Parameter Estimation Using Particle Filters

### 2010-09-01 doc/thesis/intro.tex
lots of fixes from the corrections

\section{\ac{fMRI}}
\ac{MRI}, is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in different molecules. First, the subject 
is brought into a large magnetic field which causes nuclear spins
to align. \ac{RF} signals may
then be used to excite nuclear spin away from the base alignment. 
As the nuclei precess back to the alignment of the magnetic
field, they emit detectable \ac{RF} signals. Conveniently, the
nuclear spins return their original state at different
rates, called the \ac{T1} relaxation time, depending on the material excited.
Additionally, the
coherence of the spins also decays differently (and roughly an order of 
magnitude faster
than \ac{T1} relaxation) based on the properties of the region.
This gives two primary methods of contrasting substances,
which form the basis of \ac{T1} and \ac{T2} weighted images. Additionally, 
dephasing occurs at two different rates, the \ac{T2} relaxation time,
which is unrecoverable, and \ac{T2}$^*$ relaxation, which is
much faster, but possible to recover from via special \ac{RF} signals.
\ac{T1} relaxation times are typically on the order of seconds if 
a sufficiently strong excitation is applied, whereas \ac{T2}$^*$ relaxation
times are usually less than 100ms. 
In order to rapidly acquire entire brain images, as done in Functional 
\ac{MRI}, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single \ac{T1} relaxation period. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the \ac{SNR} is very low
in \ac{EPI}.

Increasing the spatial resolution of \ac{EPI} necessarily 
requires more time or faster magnetic field switching. Increasing
switching rates can result in
more artifacts and lower signal to noise ratios. The result is
that at best \ac{fMRI} is capable of 1 second temporal resolution. 
The signal is also diluted because each voxel contains
a variety of neurons, capillaries and veins. 
Thus, the \ac{fMRI} signal, which is sensitive to the chemical composition of 
materials, is the average signal from various types of tissue
in addition to the blood. As mentioned in \autoref{sec:Introduction Overview},
and explored in depth in \autoref{sec:BOLD Physiology},
the usefulness of \ac{fMRI} comes from discerning changes in 
\ac{dHb}/\ac{O2Hb}. Therefore, it is necessary to assume
that the only chemical changes will be in
capillary beds feeding neurons. In practice this may not be the case, for
instance near large veins, and it may explain some of the
noise seen in \ac{fMRI} imaging (see \autoref{sec:Introduction Noise}). 
Because \ac{MRI} is unitless and certain
areas will have a higher base \ac{MR} signal, all \ac{fMRI} studies deal with
percent change from the base signal; rather than raw values.

\section{\ac{BOLD} Physiology}
\label{sec:BOLD Physiology}
It is well known that the two types of \ac{Hb} act as a contrast agents in 
\ac{EPI} imaging \cite{Buxton1998, WEISSKOFF1994, Ogawa}, however the connection
between \ac{dHb}/\ac{O2Hb} and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase \ac{dHb}, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow, accomplished by loosening 
capillary beds, precedes increased outflow, driving increased 
blood storage.
Since the local \ac{MR} signal depends on the ratio of \ac{dHb} to \ac{O2Hb},
increased blood volume affects this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model \cite{Buxton1998} and windkessel
model \cite{Mandeville1999}. These works derive, from first principals,
the changes in \ac{dHb} ratio and volume of capillaries given an inflow waveform.
These were the first two attempts to quantitatively account for the shape of the 
\ac{BOLD} signal as a consequence of the lag between the \ac{CBV}
and the inward \ac{CBF}.

@@

Because \autoref{eq:bold} is not capable of producing such a prolonged undershoot,
additional factors must be present. Two prominent theories exist to explain the post 
stimulus undershoot.  Recall
that a lower than base signal means that there is an increased \ac{dHb}
content in the voxel. The first and simplest explanation is that the post-stimulus
undershoot is caused by a prolonged increase in \ac{CMRO2} after \ac{CBV} and \ac{CBF}
have returned to their base levels. This theory is justified by 
studies that show \ac{CBV} and \ac{CBF} returning to the baseline before the \ac{BOLD} signal
\cite{Frahm2008, Donahue2009, Buxton2004, Lu2004, Shen2008}. 
Unfortunately, because of limitations on \ac{fMRI} and in vivo
\ac{CBV}/\ac{CBF} measurement techniques it is difficult to isolate whether \ac{CBF} and
\ac{CBV} truly have returned to their baseline. Other studies indicate
that there can be a prolonged supernormal \ac{CBV}, although none of these papers completely
rule out the possibility of increased \ac{CMRO2}  \cite{Mandeville1999a,
Behzadi2005, Chen2009a}. The discrepancies may in part
be explained by a spatial dependence in the post-stimulus undershoot; described
by Yacoub et al. \cite{Yacoub2006}. In Chen et al. 
a compelling case is made that most of the post stimulus undershoot can be 
explained by combination of a prolonged \ac{CBV} increase, and a prolonged \ac{CBF} 
undershoot \cite{Chen2009}. Additionally it was proposed that
the previous measurements showing a quick recovery of \ac{CBV} 
were in fact showing a return to baseline by arterial \ac{CBV} (which
has little effect on the \ac{BOLD} signal).

@@

\section{Properties of the \ac{BOLD} Model}
\label{sec:BOLD Analysis}
Since the first complete \ac{BOLD} model was proposed by Friston et al.,  
several studies have analyzed its properties \cite{Friston2002}.  
The most important property is that the system is dissipative, and given
enough time will converge to a constant value. This is found simply by
analyzing the eigenvalues of the state equation Jacobian, 
\cite{Deneux2006, Hu2009}. The steady state of the Balloon
model equations gives:

@@

%note to self, friston2002b's parameters are from a picture
\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & Friston et al. \cite{Friston2002} & Jonston et al. \cite{Johnston2008} 
    & Vakorin et al. \cite{Vakorin2007} & Deneux et al. \cite{Deneux2006}\\
\hline
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .045^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
\hline
\end{tabular}
\caption{Parameters found in various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

>>>>

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{images/bold}
\caption[Diagram of BOLD Signal]{A) End of Pre-Stimulus Undershoot, $< 2s$ typical,
B) End of Stimulus, C) Input Bloodflow Drops from Supernormal to Subnormal,
$5 s$ after B typical, D) Beginning of Post-stimulus Undershoot, $12 s$ after B
typical, E) Full return to normal, between $20-30s$ after B. \cite{Behzadi2005}}
\end{figure}

\section{\ac{fMRI}}
\ac{MRI}, is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in different molecules. The technique is
well established and documented in various works \cite{Bushberg2002}. 
First, the subject 
is brought into a large magnetic field which causes nuclear spins
to align. \ac{RF} signals may
then be used to excite nuclear spin away from the base alignment. 
As the nuclei precess back to the alignment of the magnetic
field, they emit detectable \ac{RF} signals. Conveniently, the
nuclear spins return their original state at different
rates, called the \ac{T1} relaxation time, depending on the material excited.
Additionally, the
coherence of the spins also decays differently (and roughly an order of 
magnitude faster
than \ac{T1} relaxation) based on the properties of the region 
\cite{Bushberg2002}.
This gives two primary methods of contrasting substances,
which form the basis of \ac{T1} and \ac{T2} weighted images. Additionally, 
dephasing occurs at two different rates, the \ac{T2} relaxation time,
which is unrecoverable, and \ac{T2}$^*$ relaxation, which is
much faster, but possible to recover from via special \ac{RF} signals.
\ac{T1} relaxation times are typically on the order of seconds if 
a sufficiently strong excitation is applied, whereas \ac{T2}$^*$ relaxation
times are usually less than 100ms. 
In order to rapidly acquire entire brain images, as done in Functional 
\ac{MRI}, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single \ac{T1} relaxation period
\cite{Obata2004}. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the \ac{SNR} is very low
in \ac{EPI} \cite{Smith2007}.

Increasing the spatial resolution of \ac{EPI} necessarily 
requires more time or faster magnetic field switching. Increasing
switching rates can result in
more artifacts and lower signal to noise ratios. The result is
that at best \ac{fMRI} is capable of 1 second temporal resolution
\cite{Ladd2010, Poser2010}. 
The signal is also diluted because each voxel contains
a variety of neurons, capillaries and veins \cite{Obata2004}. 
Thus, the \ac{fMRI} signal, which is sensitive to the chemical composition of 
materials, is the average signal from various types of tissue
in addition to the blood. As mentioned in \autoref{sec:Introduction Overview},
and explored in depth in \autoref{sec:BOLD Physiology},
the usefulness of \ac{fMRI} comes from discerning changes in 
\ac{dHb}/\ac{O2Hb} \cite{Ogawa, Obata2004}. Therefore, it is necessary to assume
that the only chemical changes will be in
capillary beds feeding neurons. In practice this may not be the case, for
instance near large veins, and it may explain some of the
noise seen in \ac{fMRI} imaging (see \autoref{sec:Introduction Noise}). 
Because \ac{MRI} is unitless and certain
areas will have a higher base \ac{MR} signal, all \ac{fMRI} studies deal with
percent change from the base signal; rather than raw values.

\section{BOLD Physiology}
\label{sec:BOLD Physiology}
It is well known that the two types of \ac{Hb} act as contrast agents in 
\ac{EPI} imaging \cite{Buxton1998, WEISSKOFF1994, Ogawa}, however the connection
between \ac{dHb}/\ac{O2Hb} and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase \ac{dHb}, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow, accomplished by loosening 
capillary beds, precedes increased outflow, driving increased 
blood storage.
Since the local \ac{MR} signal depends on the ratio of \ac{dHb} to \ac{O2Hb},
increased blood volume affects this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model \cite{Buxton1998} and windkessel
model \cite{Mandeville1999}. These works derive, from first principals,
the changes in \ac{dHb} ratio and volume of capillaries given an inflow waveform.
These were the first two attempts to quantitatively account for the shape of the 
\ac{BOLD} signal as a consequence of the lag between the \ac{CBV}
and the inward \ac{CBF}.

@@

Because \autoref{eq:bold} is not capable of producing such a prolonged undershoot,
additional factors must be present. Two prominent theories exist to explain the post 
stimulus undershoot.  Recall
that a lower than base signal means that there is an increased \ac{dHb}
content in the voxel. The first and simplest explanation is that the post-stimulus
undershoot is caused by a prolonged increase in \ac{CMRO2} after \ac{CBV} and \ac{CBF}
have returned to their base levels. This theory is justified by 
studies that show \ac{CBV} and \ac{CBF} returning to the baseline before the \ac{BOLD} signal
\cite{Frahm2008, Donahue2009, Buxton2004, Lu2004, Shen2008}. 
Unfortunately, because of limitations on \ac{fMRI} and in vivo
\ac{CBV}/\ac{CBF} measurement techniques it is difficult to isolate whether \ac{CBF} and
\ac{CBV} truly have returned to their baseline. Other studies indicate
that there can be a prolonged supernormal \ac{CBV}, although none of these papers completely
rule out the possibility of increased \ac{CMRO2}  \cite{Mandeville1999a,
Behzadi2005, Chen2009}. The discrepancies may in part
be explained by a spatial dependence in the post-stimulus undershoot; described
by Yacoub et al. \cite{Yacoub2006}. In Chen et al. 
a compelling case is made that most of the post stimulus undershoot can be 
explained by combination of a prolonged \ac{CBV} increase, and a prolonged \ac{CBF} 
undershoot \cite{Chen2009}. Additionally it was proposed that
the previous measurements showing a quick recovery of \ac{CBV} 
were in fact showing a return to baseline by arterial \ac{CBV} (which
has little effect on the \ac{BOLD} signal).

@@

\section{Properties of the \ac{BOLD} Model}
\label{sec:BOLD Analysis}
Since the first complete \ac{BOLD} model was proposed by Friston et al. 
\cite{Friston2000},  
several studies have analyzed its properties.  
The most important property is that the system is dissipative, and given
enough time will converge to a constant value. This is found simply by
analyzing the eigenvalues of the state equation Jacobian, 
\cite{Deneux2006, Hu2009}. The steady state of the Balloon
model equations gives:

@@

%note to self, friston2002b's parameters are from a picture
\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & Friston et al. \cite{Friston2000} & Jonston et al. \cite{Johnston2007} 
    & Vakorin et al. \cite{Vakorin2007} & Deneux et al. \cite{Deneux2006}\\
\hline
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .045^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
\hline
\end{tabular}
\caption{Parameters found in various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2000}
where not explicitly stated}
\label{tab:Params} 
\end{table}

### 2010-09-01 doc/thesis/particlefilters.tex
small typos changed
lots of fixes from the corrections

\chapter{Particle Filters}
\label{sec:Particle Filter}
\section{Introduction}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a timeseries of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. The idea behind particle filters is
similar to Kalman Filters; however, unlike Kalman Filters,
distributions are stored as an empircal distribution rather than the
as the first two moments of a Gaussian. Thus, particle filters are 
preferred when the model is nonlinear, and non-gaussian. This section
is largely based on Arulampalam et al. and Thrun et al. \cite{Arulampalam2002a, Thrun2005}.

@@

Suppose a set or stream of measurements at discrete times are given, 
$\{Y_k, k = 1, 2, 3, ... K\}$, where $K$ may be infinite. 
Because $k$ is a discrete time, let $t_k$ define the continuous
time of $k$.
Suppose also that there is a hidden set of state variables,
$X(t)$ that drives the value of $Y(t)$. Throughout this section
with $X_k = X(t_k)$. The goal of the particle filter is to estimate the 
distribution of the
true parameters $\Theta$ that dictates the movement of $X(t)$.
The model also permits random motion in $X(t)$, so the 
particle filter also estimates the distribution of $X(t)$.
The only difference between the members of parameter vector
$\Theta$ and those of $X(t)$ is that the memebers of
$\Theta$ have no known update equation. Members of both vectors
are permitted to have some noise, although this
may not be explicitly stated in the model. The generic, continuous, nonlinear
system definition is shown in \autoref{eq:GenericNonlinear}.

@@

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the BOLD signal, the period varies too much
for the system to be considered varying with time. 
Next, its assumes that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions allow for a simplified version of the
state space equations:

@@

To begin with, the particle filter must be given a prior distribution, from
which the initial $N_p$ particles are drawn. A particle contains a weight
as well as an estimate of $X_k$, which as previously mentioned, contains every
variable needed to run the model. Then the prior is generated from a 
given distribution, $\alpha(X)$, by:

@@

Where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF. 
Note that any exponents will be explicitly labeled as such, to avoid confusion with
the particle numbering scheme.

Therefore, after the particle have been generated they should approximate $\alpha(X)$:

\begin{equation}
\alpha(X) \approx P(x_0) = \sum_{i=0}^{N_p} w^i\delta(X - x^i_0 ) dx
\end{equation}
Where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ (the Kronecker delta function).

If a flat prior is 
preferred, then each particle's weight could be scaled to the reciprocal of the
density at the particle: 
\begin{equation}
w^i = \frac{1}{\alpha(x^i_0)}
\end{equation}
Whether or not to flatten the prior is a design decision. The reason this might 
be preferred over a direct
uniform distribution is that the distribution width will inherently scale 
for increased particle counts although some distributions
flatten out better than others. Either way, $\alpha(X)$ \emph{must} be
wide enough to incorporate any posterior that arises. If the prior is
not sufficiently dense, the particle filter may be able to compensate, if it is
not sufficiently wide the particle filter won't converge.

@@

Whenever a measurement becomes available it permits refinement of the
posterior density.
This process of incorporating new data is called sequential importance sampling,
and eventually allows convergence. The weight is defined as
\begin{equation}
w^i_k \propto \frac{P(x^i_{0:k} | y_{0:k})}{q(x^i_{0:k} | y_{0:k})}
\label{eq:weightfunc}
\end{equation}
where $q$ is called an \emph{importance density}. The importance density
is the density of the points, thus by dividing by this value, the weight
should not depend on the location of the estimation points, but rather
only on $P(x^i_{0:k} | y_{0:k})$, the probability of that particle
being correct given all the measurements up to time $k$. 
Note that if there is a far off peak in
the posterior that $q$ does not have support points in, there will 
be quantization errors, and that part of the density cannot be modeled. This is why
it is absolutely necessary that $q$ fully covers $P(x^i_{0:k} | y_{0:k})$.

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$. For practical applications though it is a helpful assumption.
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
The benefit of this choice for importance density is that an approximation for
$P(x_k | x_{k-1})$ is freely available: its simply the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen in \autoref{eq:weightevolve}.

@@

Calculating $P(x_{0:k} | y_{0:k})$ is a bit more involved. 
First, using the assumption that the distribution of $y_k$ is 
fully constrained by $x_k$, and that $x_k$ is similarly fully 
constrained by $x_{k-1}$, I make the good assumptions that:
\begin{eqnarray}
P(y_k | x_{0:k}, y_{0:k-1}) &=& P(y_k | x_k) \nonumber \\
P(x_k | x_{0:k}, y_{0:k-1}) &=& P(x_k | x_{k-1})
\label{eq:MarkovProperty}
\end{eqnarray}
These are of course just re-statements of the state equations assumed by \autoref{eq:stateass}
and \autoref{eq:measass}.

@@

Thus, by making the following simple assumptions, evolving a posterior
density  requires no knowledge of noise distribution.
\begin{enumerate}
\item $f(t, x(t), u(t)) = f(x(t), u(t))$ and $g(t, x(t), u(t)) = g(x(t))$ 
\item The prior distribution PDF, $q(x_i(0))$, covers $P(x_i(0))$
\item Markov Property: $P(x_k | x_{0:k-1}) = Pr(x_k | x_{k-1})$
\item $q(x_{0:k-1} | y_{0:k}) = q(x_{0:k-1} | y_{0:k-1})$
\end{enumerate}

>>>>

\chapter{Particle Filters}
\label{sec:Particle Filter}
\section{Introduction}
Particle filters, a type of Sequential Monte Carlo (SMC) method,
are a powerful method for estimating the posterior probability distribution
 of parameters given a timeseries of measurements. Unlike Markov 
Chain Monte Carlo (MCMC) estimation, particle filters are designed for 
time-varying random variables. The idea behind particle filters is
similar to Kalman Filters; however, unlike Kalman Filters,
distributions are stored as an empircal distribution rather than 
as the first two moments of a Gaussian. Thus, particle filters are 
preferred when the model is nonlinear, and non-gaussian. This section
is largely based on Arulampalam et al. and Thrun et al. \cite{Arulampalam2002a, Thrun2005}.

@@

Suppose a set or stream of measurements at discrete times are given, 
$\{Y_k, k = 1, 2, 3, ... K\}$, where $K$ may be infinite. 
Let $k$ denote a time index and $t_k$ define the continuous
time of index $k$.
Suppose also that there is a hidden set of state variables,
$X(t)$ that drives the value of $Y(t)$. Throughout this section
$X_k$ will identify $X(t_k)$. The goal of the particle filter is to estimate the 
distribution of the
true parameters $\Theta$ that dictates the movement of $X(t)$.
The model also permits random motion in $X(t)$, so the 
particle filter also estimates the distribution of $X(t)$.
The only difference between the members of parameter vector
$\Theta$ and those of $X(t)$ is that the members of
$\Theta$ have no known update equation. Members of both vectors
are permitted to have some noise, although this
may not be explicitly stated in the model. The generic, continuous, nonlinear
system definition is shown in \autoref{eq:GenericNonlinear}.

@@

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the BOLD signal, the period varies too much
for the system to be considered varying with time. 
Next, it was assumed that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions allow for a simplified version of the
state space equations:

@@

To begin, the particle filter must be given a prior distribution, from
which the initial $N_p$ particles are drawn. A particle contains a weight
as well as an estimate of $X_k$, which as previously mentioned, contains every
variable needed to run the model. Then the prior is generated from a 
given distribution, $\alpha(X)$, by:

@@

where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF. 
Note that any exponents will be explicitly labeled as such, to avoid confusion with
the particle numbering scheme.

Therefore, after the particles have been generated they should approximate $\alpha(X)$:

\begin{equation}
\alpha(X) \approx P(x_0) = \sum_{i=0}^{N_p} w^i\delta(X - x^i_0 ) dx
\end{equation}
where $\delta(x-x_0)$ is 1 if and only if $x = x_0$ 
(the Kronecker delta function).

If a flat prior is 
preferred, then each particle's weight could be scaled to the reciprocal of the
density at the particle: 
\begin{equation}
w^i = \frac{1}{\alpha(x^i_0)}
\end{equation}
Whether or not to flatten the prior is a design decision. The reason this might 
be preferred over a direct
uniform distribution is that the distribution width will inherently scale 
for increased particle counts although some distributions
flatten out better than others. Either way, $\alpha(X)$ must be
wide enough to incorporate any posterior that arises. If the prior is
not sufficiently dense, the particle filter may be able to compensate, if it is
not sufficiently wide the particle filter won't converge.

@@

Whenever a measurement becomes available it permits refinement of the
posterior density.
This process of incorporating new data is called sequential importance sampling,
and eventually allows convergence. The weight is defined as
\begin{equation}
w^i_k \propto \frac{P(x^i_{0:k} | y_{0:k})}{q(x^i_{0:k} | y_{0:k})}
\label{eq:weightfunc}
\end{equation}
where $q$ is called an \emph{importance density}. The importance density
is the density of the points, thus by dividing by this value, the weight
should not depend on the location of the estimation points, but rather
only on $P(x^i_{0:k} | y_{0:k})$, the probability of that particle
being correct given all the measurements up to time $k$. 
Note that if there is a distant peak in
the posterior that $q$ does not have support points in, there will 
be quantization errors, and that part of the density cannot be modeled. 
This is why it is absolutely necessary that $q$ fully covers 
$P(x^i_{0:k} | y_{0:k})$.

@@

\subsection{Calculating Weights}
To calculate the weight of a particular particle, it is necessary to 
calculate both $q(x^i_{0:k} | y_{0:k})$ and $P(x^i_{0:k} | y_{0:k})$.
Note that $q(x^i_{0:k} | y_{0:k})$ may be simplified by assuming that 
$y_k$ doesn't contain any information about $x_{k-1}$. Technically this 
could be false; since later measurements may shed light on currently hidden
changes in $x$, however in this work it is a good assumption because the
input does not immediately effect the BOLD signal. 
\begin{equation}
q(x^i_{0:k} | y_{0:k}) = q(x^i_{0:k} | y_{0:k-1})
\label{eq:QAssump}
\end{equation}
The choice of the importance density is another design decision; however
it is common to use the integrated state equations. 
Although other importance density functions exist; for the particle filter
used here, the standard importance density will be used: the model
prior.
\begin{equation}
q(x_k | x_{k-1}, y_{0:k}) =  P(x_k | x_{k-1})
\label{eq:ImportanceDensity}
\end{equation}
The benefit of this choice for importance density is that an approximation for
$P(x_k | x_{k-1})$ is freely available; its simply the set of particles propagated
forward in time using the state equations. Additionally it makes
updating weights simple, as seen in \autoref{eq:weightevolve}.

@@

Calculating $P(x_{0:k} | y_{0:k})$ is a bit more involved. 
First, using the assumption that the distribution of $y_k$ is 
fully constrained by $x_k$, and that $x_k$ is similarly fully 
constrained by $x_{k-1}$, I make the assumptions that:
\begin{eqnarray}
P(y_k | x_{0:k}, y_{0:k-1}) &=& P(y_k | x_k) \nonumber \\
P(x_k | x_{0:k}, y_{0:k-1}) &=& P(x_k | x_{k-1})
\label{eq:MarkovProperty}
\end{eqnarray}
These are of course just re-statements of the state equations assumed 
by \autoref{eq:stateass} and \autoref{eq:measass}.

@@

Thus, by making the following simple assumptions, evolving a posterior
density  requires no knowledge of the noise distribution.
\begin{enumerate}
\item $f(t, x(t), u(t)) = f(x(t), u(t))$ and $g(t, x(t), u(t)) = g(x(t))$ 
\item The prior distribution PDF, $q(x_i(0))$, covers $P(x_i(0))$
\item Markov Property: $P(x_k | x_{0:k-1}) = Pr(x_k | x_{k-1})$
\item $q(x_{0:k-1} | y_{0:k}) = q(x_{0:k-1} | y_{0:k-1})$
\end{enumerate}
